### 2022-10-22 dissertation/chapters/03-heuristics.tex
🚧 WIP - Save point (compiles)

\noindent
Many different techniques have been used to train \acp{FFNN} \cite{ref:kingma:2014}. Finding the best technique to use to train an \acs{FFNN} has been shown to be problem dependent in many cases \cite{ref:kheiri:2017}. Every technique has its characteristics, constraints, advantages and disadvantages. At the time of writing, the majority of work that is published around the training of \acp{FFNN}, involves the use of gradient-based techniques \cite{ref:nel:2021}. Gradient-based techniques are not without flaws and can, for example, yield slow convergence or get trapped in local optima \cite{ref:mingguang:2009}. Other techniques have also been used to successfully train \acp{FFNN}, including meta-heuristics such as \acf{PSO} \cite{ref:rakitianskaia:2012, ref:vanwyk:2014}, \acf{DE} \cite{ref:espinal:2011} and \acfp{GA} \cite{ref:gupta:1999}.

@@

Optimisation is the task of finding a solution to a given problem that is better than alternative solutions. Better stated by Oldewage \cite{ref:oldewage:2017}, optimisation is the task of finding values for a set of variables such that some measure of optimality is satisfied given a set of constraints. Engelbrecht \cite{ref:engelbrecht:2007} breaks optimisations problems down into three components:

@@

As stated in Chapter~\ref{chap:anns}, the training of an \acp{FFNN} is a particular type of optimisation problem, where the goal is to find the configuration of \textit{weights}, such that the \ac{FFNN} yields output that minimises some loss function. The mechanism by which the optimal weights for a \acs{FFNN} is sough out, is executed by an \index{optimisation algorithm}optimisation algorithm known as a \index{heuristic}heuristic. Discussions on the details of heuristics are presented in the following sections.

@@

The term \textit{heuristic} comes from the Latin word \textit{heuristicus} which means ``to find out or discover''. \citeauthor{ref:romanycia:1985} \cite{ref:romanycia:1985} provides a complete study on the history and origins of the term \index{heuristic}\textit{heuristic}. From their research, a proposal is made to define \index{heuristic}heuristics in the context of \ac{AI}, as any device, be it a program, rule, piece of knowledge, which is added to a problem-solving system, in expectation that on average, the performance will improve.

In the context of this dissertation, a \index{heuristic}heuristic refers to an algorithmic search technique that serves as a guide to a search process where good solutions to a optimisation problem is being sought out. Different \index{heuristic}heuristics make use of different information during the search process \cite{ref:kheiri:2017}. During the training of \acp{FFNN}, \index{heuristic}heuristics such as gradient-based \index{heuristic}heuristics make use of the derivatives obtained by evaluating the \ac{FFNN}. It can thus be said that gradient-based \index{heuristic}heuristics make use of information directly from the \textit{search space}. On the contrary, \index{heuristic}heuristics such as \index{hmeta-euristic}meta-heuristics make use of meta-information obtained as a result of evaluating the \ac{FFNN} \cite{ref:blum:2003}. The meta-information that \index{meta-heuristic}meta-heuristics make use of could include ranked-performance of a population of candidate solutions, referred to as \textit{entities}. \index{meta-heuristic}Meta-heuristics are useful when there is imperfect information about the search space \cite{ref:bianchi:2009}, and are generally less problem-specific than other classes of \index{heuristic}heuristics \cite{ref:blum:2003}. This disseration takes a particular interest in gradient-based \index{heuristic}heuristics and \index{meta-heuristic}meta-heuristics. Each of these are presented in detail in the following sections.

@@

Gradient-based \index{heuristic}heuristics are optimisation techniques that make use of derivates obtained from evaluating the model being optimised. Specifically, in the context of a minimisation problem, these techniques are called \acf{GD} heuristics as they \textit{minimise} some loss function. \Acs{GD} is generally attributed to Cauchy \cite{ref:lemarechal:2012}, who first suggested it in 1847. In 1907 Hadamard \cite{ref:hadamard:1908} independently proposed a similar method.

Although \acs{GD} \index{heuristic}heuristics where not the first \index{heuristic}heuristics used to train \ac{FFNN}
\cite{ref:engelbrecht:2007}, they are certainly the most whidely used. \Acs{GD} heuristics have become increasingly popular partly due to their simplicity and low computational overhead compared to other \index{heuristic}heuristics such as \index{meta-heuristic}meta-heuristics and other second-order derivative methods such as Newtons' method. There are many variants of \acs{GD} \index{heuristic}heuristics, however, they all fundamentally apply the same generic \acs{GD} framework called \acl{BP}.

@@

Chapter~\ref{chap:anns} introduced the reader to \index {supervised learning}supervised learning and presented a number of loss functions. In the context of supervised learning, loss functions produce a scalar value $\epsilon$, that represents the error between the output of the \ac{FFNN} and the desired output. When using \acs{GD} to train \acp{FFNN}, the \index{loss function}loss function is used to adjust the weights of the \ac{FFNN} in order to minimise the error \cite{ref:engelbrecht:2007}. \citeauthor{ref:engelbrecht:2007} \cite{ref:engelbrecht:2007} states that training of \acp{FFNN} using \acs{GD}, is done by calculating the gradient of $\epsilon$ in \textit{weight-space}, and then moving the weight vector along the negative gradient. An illustration of \acs{GD} is given in Figure~\ref{fig:heuristics:gd:gd_illustration}.

\begin{figure}[htbp]
      \centering
      \includegraphics[width=0.85\textwidth]{images/gradient_descent.pdf}
      \caption{An illustration of \ac{GD} over various timesteps showing the minimisation of the error with regards to weight value.}
      \label{fig:heuristics:gd:gd_illustration}
\end{figure}

\noindent
The update step for \ac{GD} can be formulated as is shown in Equation~\eqref{eq:heuristics:gd:update_step_part_1} below.

@@

\noindent
In the context of training \textit{shallow} \acp{FFNN} using \index{supervised learning}supervised learning, the error signal is propagated backwards in the network from the output layer, through the hidden layer. The algorithm that propagates the error signal backwardds is known as \acf{BP}. \Acs{BP} was popularised by \citeauthor{ref:werbos:1994} \cite{ref:werbos:1994}. \citeauthor{ref:nel:2021} \cite{ref:nel:2021} states that \acs{BP} provides a procedure for updating the network weights, layer by layer, by evaluating the derivatives of the error function $\epsilon$ with respect to the weights at each layer. \citeauthor{ref:engelbrecht:2007} \cite{ref:engelbrecht:2007}describes the \acs{BP} process in two steps:

\begin{itemize}
      \item \textbf{Feedforward Pass}: During this phase the output values of the \acs{FFNN} is calculcated for each training pattern.
      \item \textbf{Backward Propagation}: During this phase the error signal that is calculated as is shown above in Equations~\eqref{eq:heuristics:gd:update_step_part_1}-~\eqref{eq:heuristics:gd:update_step_part_3} is propagated backwards from the output layer, through the hidden layer, to the input layer of the \ac{FFNN}. Weights are then adjusted as functions of the backpropagated error signal.
\end{itemize}

@@

The pseudo-code implementation for the \ac{SGD} algorithm is taken from \cite{ref:engelbrecht:2007} and is presented in Algorithm~\ref{algo:heuristics:gd:sgd} below.

\begin{algorithm}[H]
      \caption{The pseudo code algorithm for the \ac{SGD} heuristic.}
      \label{algo:heuristics:gd:sgd}
      \begin{algorithmic}
            \State Initialise weights, $\eta$, $\alpha$ and the number of epochs $t=0$;
            \While{stopping condition not met}
            let $\epsilon_{T} = 0$
            \For{each training pattern p}
            \State Do the feedforward pass phase to calculate $y_{j,p}$ ($\forall j = 1, \dots, J$) and $o_{k,p}$ ($\forall k = 1, \dots, K$);
            \State Compute the output error signals $\delta_{o_{k,p}}$ and the hidden layer error signals $\delta_{y_{j,p}}$;
            \State Adjust the weights $w_{kj}$ and $v_{ji}$ (backpropagation of errors);
            \State $\epsilon_{T}$ += $[\epsilon_{p} = \sum^{K}_{k=1}(t_{k,p} - o_{k,p})^{2}]$;
            \EndFor
            \State $t = t + 1$;
            \EndWhile
      \end{algorithmic}
\end{algorithm}

In order to provide a simplified frame of reference from which the gradient-based heuristics are compared, consider a simplification of the update-step for the weights of the \ac{FFNN} as is done using \ac{SGD} below.

@@

In Equation~\ref{eq:heuristics:gd:sgd} $w$ refers to the weights of the \ac{FFNN}, $\eta$ refers to the learning rate as before and $\nabla_{w}\mathcal{E}(w)$ refers to the gradient of the loss/cost function w.r.t the output of the \ac{FFNN} as a result of its weights. Note that all subscripts w.r.t. layers and training patterns are omitted for convenience.

So far detailed mathematical descriptions have been provided to illustrate the concept of \ac{GD} and how it applies to the \ac{SGD} heuristic as applied to \ac{FFNN} training. It was mentioned that \ac{SGD} was one of the first widely used heuristics to train \ac{FFNN}, however, it is not without flaws. With stochastic learning, only one training pattern is presented at each iteration. As such, weight updates are done with high variance and noise \cite{ref:ruder:2016}. An illustration of the the fluctuations caused by \ac{SGD} during training is given in Figure~\ref{fig:heuristics:gd:sgd}.

\begin{figure}[htbp]
      \includegraphics[width=\textwidth]{images/sgd.png}
      \caption{An illustration of \ac{SGD} fluctuations during training as taken from \cite{ref:sgd:2006}.}
      \label{fig:heuristics:gd:sgd}
\end{figure}

With batch learning, all the training patterns are presented at once and the \ac{FFNN} converges to the minimum of the loss/cost function w.r.t $w$. However, this is computationally expensive and intractable to implement in many situations. While \ac{SGD} is able to jump out of local minima into new, potentially better local minima \cite{ref:ruder:2016}, it does at first this seems advantageous. However, this complicates convergence to the exact minima as \ac{SGD} can potentially keep overshooting better minima. It is shown that a slow decrease in the learning rate $\eta$ will lead to the same convergence behaviour as with batch learning, almost certainly converging to local or global minima for non-convex and convex optimisation respectively. A compromise to this problem is to include mini-batches of training patterns where a small number of training patterns are presented to the \ac{FFNN} at once. This means that the input patterns from mini-batches are just approximations of the total population of training patterns. According to \citeauthor{ref:ruder:2016} \cite{ref:ruder:2016} this has two main advantages

@@

      \item Learning rate schedules \cite{ref:robbins:1951} can be introduced to dynamically change the learning rate throughout the training process, however, these schedules and their parameters have to be defined a priori and are thus problem specific \cite{ref:darken:1992}.

      \item The learning rate that has been introduced so far is applied to all weight updates. If the training data is sparse and the features have very different frequencies an equal update to all weights is inefficient. Larger weight updates are required for less frequently occurring features. This problem eludes to the credit assignment problem \cite{ref:rumelhart:1986} common to these \ac{GD} variants.

      \item It is difficult to avoid getting trapped in local minima, especially for highly non-convex loss/cost functions used for \ac{ANN} training. \cite{ref:dauphin:2014} argues that this difficulty arises not from local minima, but from sadle points. Sadle points are points at which one dimensions slopes upwards while another dimension slopes downwards. \cite{ref:ruder:2016} mentions that these sadle points are usually surrounded by plateaus of the same error, leading to gradients that are close to zero in all dimensions.
\end{itemize}

@@

Research shows that \ac{SGD} has difficulty navigating ravines i.e. areas where the surface curves much more steeply in one dimension than in another \cite{ref:sutton:1986}. These ravines are common around local minima. As such, \ac{SGD} is shown to oscillate across the slopes of the ravine while only making minor progress towards the local minima. An illustration of this is given in Figure~\ref{fig:heuristics:gd:sgd_with_and_without_momentum}.

\begin{figure}[htbp]
      \includegraphics[width=\textwidth]{images/sgd_with_and_without_momentum.png}
      \caption{An illustration of \ac{SGD} with and without momentum taken from \cite{ref:du:2019}.}
      \label{fig:heuristics:gd:sgd_with_and_without_momentum}
\end{figure}

Momentum \cite{ref:qian:1999} is a variant of the \ac{SGD} that helps accelerate \ac{SGD} in the relevant direction, dampening oscillations. \cite{ref:ruder:2016} mentions that this is done by adding a fraction $\alpha$ of the weight update vector of the previous timestep to the current weight update vector. The accumulation of momentum is presented in Equation~\ref{eq:heuristics:gd:momentum_part_1} while the update step is then amended in Equation~\ref{eq:heuristics:gd:momentum_part_2} below.

@@

By redefining the \ac{SGD} update steps as shown above, the \ac{Momentum} heuristic allows for the increase of momentum for dimensions whose gradients point in the same direction, while simultaneously reducing momentum for dimensions whose gradients change direction, leading to faster convergence and less oscillation. The momentum term $\alpha$ is usually set to 0.9 \cite{ref:engelbrecht:2007}\cite{ref:ruder:2016}.

@@

\Acl{NAG} is a variant of the \ac{Momentum} heuristics developed by Ilya Sutskever \cite{ref:sutskever:2013-2} and is inspired from Yurii Nesterov's \cite{ref:nesterov:1983} work on optimising convex functions. \ac{NAG} provides an improvement to the momentum accumulation term by providing a look-ahead term that better refines the weight update step. In the \ac{NAG} heuristic, the gradient is not calculated w.r.t the current weights, but rather w.r.t the approximate future positions of the weights \cite{ref:ruder:2016}. An illustration of this is taken from Geoffrey Hinton's lecture on mini-batch \ac{GD} \cite{ref:hinton:2012} and is presented in Figure~\ref{fig:heuristics:gd:nag} below.

\begin{figure}[htbp]
      \includegraphics[width=\textwidth]{images/nag.png}
      \caption{An illustration of the weight update vector for \ac{NAG} taken from \cite{ref:hinton:2012}.}
      \label{fig:heuristics:gd:nag}
\end{figure}

\citeauthor{ref:ruder:2016} \cite{ref:ruder:2016} explains the update step as is done by \ac{NAG} is follows. \ac{Momentum} first computes the current gradient. This is seen by the small blue vector presented in Figure~\ref{fig:heuristics:gd:nag}. A large step is taken in the direction of the updated accumulated gradient, presented by the big blue vector. \ac{NAG} then first takes a big step in the direction of the previous accumulated gradient presented by the brown vector. At this point the gradient is measured and \ac{NAG} then makes a correction (red vector), which results in the complete NAG update (green vector). By anticipating approximate future positions of the weights, the weight update step as defined by \ac{NAG} controls the optimisation process from from going too fast and results in increased responsiveness. The \ac{NAG} weight vector update steps are given in Equations~\ref{eq:heuristics:gd:nag_part_1} and~\ref{eq:heuristics:gd:nag_part_2} below.

@@

\Acl{Adagrad} is an adaptation of \ac{SGD} which implements a learning rate for every parameter in the weight vector and is developed by Duchi et al. \cite{ref:duchi:2011}. \citeauthor{ref:ruder:2016} \cite{ref:ruder:2016} mentions that \Ac{Adagrad} adapts the learning rate to the weights, performing small updates (i.e. low learning rates) for weights associated with frequently occurring features and larger updates (i.e. high learning rates) for weights associated with infrequent features. For this reason \ac{Adagrad} is well suited for dealing with situations with very sparse training data.

In the \ac{GD} variants presented thus far, the weight updates have been applied to all weights at once using the same learning rate $\eta$ for all dimensions of the weight update vector. \Ac{Adagrad} uses a different learning rate for every weight $w_{i}$ at every timestep $t$. The weight update step as implemented by \ac{Adagrad} is presented in Equation~\ref{eq:heuristics:gd:adagrad_part_1} below.

@@

where $w_{i}$ is the $i$-th dimension of the weight vector, $G_{ii} \in \mathbb{R}^{d \times d}$ is a diagonal matrix where each diagonal element $i,i$ is the sum of the squared of the gradients w.r.t. $w_{i}$ up to timestep $t$ and $\epsilon$ is a smoothing term that avoids division by zero, usually set in the order of $1 \times 10^{-8}$ \cite{ref:ruder:2016}. Since $G_{t}$ contains the sum of the squares of the gradients with respect to all parameters $w$ along its diagonal, the weight update step can be vectorised and updated using the matrix-vector product between $G_{t}$ and $\nabla_{w_{i}}\mathcal{E}(w_{i})$. This is presented in Equation~\ref{eq:heuristics:gd:adagrad_part_2} below.

@@

In Section~\ref{sec:heuristics:adagrad} it was mentioned that one of the main problems of \ac{Adagrad} is that it has an aggressively monotonically decreasing learning rate because of the accumulation of the sum of squared gradients in its denominator. Zeiler \cite{ref:zeiler:2012} presents an improvement on \ac{Adagrad} that eliminates this problem. \citeauthor{ref:ruder:2016} \cite{ref:ruder:2016} mentions that instead of accumulating all past squared gradients as in the case of \ac{Adagrad}, \ac{Adadelta} restricts the window of accumulation to a window with a fixed size $W$. However, storing $W$ previous squared gradients is very inefficient. Instead, the accumulation of the sum of gradients over $W$ steps is defined recursively as a decaying average of all past squared gradients. The moving average $E[g^{2}]_{t}$ at timestep $t$ then depends only a fraction $\alpha$ of the previous average and current gradient, similar to the update step for \ac{Momentum}. In order to simplify the notation of update steps, let $g = \nabla_{w_{i}}\mathcal{E}(w_{t})$. The moving average update step for $E[g^{2}]_{t}$ is then presented in Equation~\ref{eq:heuristics:gd:adadelta_part_1} below.

@@

Since the denominator is the \ac{RMS} error criterion of the gradient, the criteria short-hand notation is used as follows.

@@

\citeauthor{ref:ruder:2016}\cite{ref:ruder:2016} highlights that the original authors \cite{ref:zeiler:2012} noted that the units of the update step as presented in Equation~\ref{eq:heuristics:gd:adadelta_part_2} do not match i.e. the update vector should have the same hypothetical units as the weight vector. To realize this, another exponentially decaying average is defined in terms of squared weight updates as is presented in Equation~\ref{eq:heuristics:gd:adadelta_part_6} below.

>>>>

\noindent
Many different techniques have been used to train \acp{FFNN}~\cite{ref:kingma:2014}. Finding the best technique to use to train an \acs{FFNN} has been shown to be problem dependent in many cases~\cite{ref:kheiri:2017}. Every technique has its characteristics, constraints, advantages and disadvantages. At the time of writing, the majority of work that is published around the training of \acp{FFNN}, involves the use of gradient-based techniques~\cite{ref:nel:2021}. Gradient-based techniques are not without flaws and can, for example, yield slow convergence or get trapped in local optima~\cite{ref:mingguang:2009}. Other techniques have also been used to successfully train \acp{FFNN}, including meta-heuristics such as \acf{PSO}~\cite{ref:rakitianskaia:2012, ref:vanwyk:2014}, \acf{DE}~\cite{ref:espinal:2011} and \acfp{GA}~\cite{ref:gupta:1999}.

@@

Optimisation is the task of finding a solution to a given problem that is better than alternative solutions. Better stated by Oldewage~\cite{ref:oldewage:2017}, optimisation is the task of finding values for a set of variables such that some measure of optimality is satisfied given a set of constraints. Engelbrecht~\cite{ref:engelbrecht:2007} breaks optimisations problems down into three components:

@@

As stated in Chapter~\ref{chap:anns}, the training of an \acp{FFNN} is a particular type of optimisation problem, where the goal is to find the configuration of \textit{weights}, such that the \acs{FFNN} yields output that minimises some loss function. The mechanism by which the optimal weights for a \acs{FFNN} is sough out, is executed by an \index{optimisation algorithm}optimisation algorithm known as a \index{heuristic}heuristic. Discussions on the details of heuristics are presented in the following sections.

@@

The term \textit{heuristic} comes from the Latin word \textit{heuristicus} which means ``to find out or discover''.~\citeauthor{ref:romanycia:1985}~\cite{ref:romanycia:1985} provides a complete study on the history and origins of the term \index{heuristic}\textit{heuristic}. From their research, a proposal is made to define \index{heuristic}heuristics in the context of \acf{AI}, as any device, be it a program, rule, piece of knowledge, which is added to a problem-solving system, in expectation that on average, the performance will improve.

In the context of this dissertation, a \index{heuristic}heuristic refers to an algorithmic search technique that serves as a guide to a search process where good solutions to a optimisation problem is being sought out. Different \index{heuristic}heuristics make use of different information during the search process~\cite{ref:kheiri:2017}. During the training of \acp{FFNN}, \index{heuristic}heuristics such as gradient-based \index{heuristic}heuristics make use of the derivatives obtained by evaluating the \acs{FFNN}. It can thus be said that gradient-based \index{heuristic}heuristics make use of information directly from the \textit{search space}. On the contrary, \index{heuristic}heuristics such as \index{hmeta-euristic}meta-heuristics make use of meta-information obtained as a result of evaluating the \acs{FFNN}~\cite{ref:blum:2003}. The meta-information that \index{meta-heuristic}meta-heuristics make use of could include ranked-performance of a population of candidate solutions, referred to as \textit{entities}. \index{meta-heuristic}Meta-heuristics are useful when there is imperfect information about the search space~\cite{ref:bianchi:2009}, and are generally less problem-specific than other classes of \index{heuristic}heuristics~\cite{ref:blum:2003}. This disseration takes a particular interest in gradient-based \index{heuristic}heuristics and \index{meta-heuristic}meta-heuristics. Each of these are presented in detail in the following sections.

@@

Gradient-based \index{heuristic}heuristics are optimisation techniques that make use of derivates obtained from evaluating the model being optimised. Specifically, in the context of a minimisation problem, these techniques are called \acf{GD} heuristics as they \textit{minimise} some loss function. \Acs{GD} is generally attributed to Cauchy~\cite{ref:lemarechal:2012}, who first suggested it in 1847. In 1907 Hadamard~\cite{ref:hadamard:1908} independently proposed a similar method.

Although \acs{GD} \index{heuristic}heuristics where not the first \index{heuristic}heuristics used to train \acp{FFNN}
\cite{ref:engelbrecht:2007}, they are certainly the most whidely used. \Acs{GD} heuristics have become increasingly popular partly due to their simplicity and low computational overhead compared to other \index{heuristic}heuristics such as \index{meta-heuristic}meta-heuristics and other second-order derivative methods such as Newtons' method. There are many variants of \acs{GD} \index{heuristic}heuristics, however, they all fundamentally apply the same generic \acs{GD} framework called \acl{BP}.

@@

Chapter~\ref{chap:anns} introduced the reader to \index {supervised learning}supervised learning and presented a number of loss functions. In the context of supervised learning, loss functions produce a scalar value $\epsilon$, that represents the error between the output of the \acs{FFNN} and the desired output. When using \acs{GD} to train \acp{FFNN}, the \index{loss function}loss function is used to adjust the weights of the \acs{FFNN} in order to minimise the error~\cite{ref:engelbrecht:2007}.~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} states that training of \acp{FFNN} using \acs{GD}, is done by calculating the gradient of $\epsilon$ in \textit{weight-space}, and then moving the weight vector along the negative gradient. An illustration of \acs{GD} is given in Figure~\ref{fig:heuristics:gd:gd_illustration}.

\begin{figure}[htbp]
      \centering
      \includegraphics[width=0.85\textwidth]{images/gradient_descent.pdf}
      \caption{An illustration of \acs{GD} over various timesteps showing the minimisation of the error with regards to weight value.}
      \label{fig:heuristics:gd:gd_illustration}
\end{figure}

\noindent
The update step for \acs{GD} can be formulated as is shown in Equation~\eqref{eq:heuristics:gd:update_step_part_1} below.

@@

\noindent
In the context of training \textit{shallow} \acp{FFNN} using \index{supervised learning}supervised learning, the error signal is propagated backwards in the network from the output layer, through the hidden layer. The algorithm that propagates the error signal backwardds is known as \acf{BP}. \Acs{BP} was popularised by~\citeauthor{ref:werbos:1994}~\cite{ref:werbos:1994}.~\citeauthor{ref:nel:2021}~\cite{ref:nel:2021} states that \acs{BP} provides a procedure for updating the network weights, layer by layer, by evaluating the derivatives of the error function $\epsilon$ with respect to the weights at each layer.~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007}describes the \acs{BP} process in two steps:

\begin{itemize}
      \item \textbf{Feedforward Pass}: During this phase the output values of the \acs{FFNN} is calculcated for each training pattern.
      \item \textbf{Backward Propagation}: During this phase the error signal that is calculated as is shown above in Equations~\eqref{eq:heuristics:gd:update_step_part_1}-~\eqref{eq:heuristics:gd:update_step_part_3} is propagated backwards from the output layer, through the hidden layer, to the input layer of the \acs{FFNN}. Weights are then adjusted as functions of the backpropagated error signal.
\end{itemize}

@@

The pseudo-code implementation for the \acs{SGD} algorithm is taken from~\cite{ref:engelbrecht:2007} and is presented in Algorithm~\ref{algo:heuristics:gd:sgd} below.

\begin{algorithm}[H]
      \caption{The pseudo code algorithm for the \acs{SGD} heuristic.}
      \label{algo:heuristics:gd:sgd}
      \begin{algorithmic}
            \State Initialise weights, $\eta$, $\alpha$ and the number of epochs $t=0$;
            \While{stopping condition not met}
            let $\epsilon_{T} = 0$
            \For{each training pattern p}
            \State Do the feedforward pass phase to calculate $y_{j,p}$ ($\forall j = 1, \dots, J$) and $o_{k,p}$ ($\forall k = 1, \dots, K$);
            \State Compute the output error signals $\delta_{o_{k,p}}$ and the hidden layer error signals $\delta_{y_{j,p}}$;
            \State Adjust the weights $w_{kj}$ and $v_{ji}$ (backpropagation of errors);
            \State $\epsilon_{T}$ += $[\epsilon_{p} = \sum^{K}_{k=1}(t_{k,p} - o_{k,p})^{2}]$;
            \EndFor
            \State $t = t + 1$;
            \EndWhile
      \end{algorithmic}
\end{algorithm}

In order to provide a simplified frame of reference from which the gradient-based heuristics are compared, consider a simplification of the update-step for the weights of the \acs{FFNN} as is done using \ac{SGD} below.

@@

In Equation~\ref{eq:heuristics:gd:sgd} $w$ refers to the weights of the \acs{FFNN}, $\eta$ refers to the learning rate as before and $\nabla_{w}\mathcal{E}(w)$ refers to the gradient of the loss/cost function w.r.t the output of the \acs{FFNN} as a result of its weights. Note that all subscripts w.r.t. layers and training patterns are omitted for convenience.

So far detailed mathematical descriptions have been provided to illustrate the concept of \ac{GD} and how it applies to the \acs{SGD} heuristic as applied to \ac{FFNN} training. It was mentioned that \ac{SGD} was one of the first widely used heuristics to train \ac{FFNN}, however, it is not without flaws. With stochastic learning, only one training pattern is presented at each iteration. As such, weight updates are done with high variance and noise~\cite{ref:ruder:2016}. An illustration of the the fluctuations caused by \ac{SGD} during training is given in Figure~\ref{fig:heuristics:gd:sgd}.

\begin{figure}[htbp]
      \includegraphics[width=\textwidth]{images/sgd.png}
      \caption{An illustration of \ac{SGD} fluctuations during training as taken from~\cite{ref:sgd:2006}.}
      \label{fig:heuristics:gd:sgd}
\end{figure}

With batch learning, all the training patterns are presented at once and the \acs{FFNN} converges to the minimum of the loss/cost function w.r.t $w$. However, this is computationally expensive and intractable to implement in many situations. While \ac{SGD} is able to jump out of local minima into new, potentially better local minima~\cite{ref:ruder:2016}, it does at first this seems advantageous. However, this complicates convergence to the exact minima as \ac{SGD} can potentially keep overshooting better minima. It is shown that a slow decrease in the learning rate $\eta$ will lead to the same convergence behaviour as with batch learning, almost certainly converging to local or global minima for non-convex and convex optimisation respectively. A compromise to this problem is to include mini-batches of training patterns where a small number of training patterns are presented to the \acs{FFNN} at once. This means that the input patterns from mini-batches are just approximations of the total population of training patterns. According to~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} this has two main advantages

@@

      \item Learning rate schedules~\cite{ref:robbins:1951} can be introduced to dynamically change the learning rate throughout the training process, however, these schedules and their parameters have to be defined a priori and are thus problem specific~\cite{ref:darken:1992}.

      \item The learning rate that has been introduced so far is applied to all weight updates. If the training data is sparse and the features have very different frequencies an equal update to all weights is inefficient. Larger weight updates are required for less frequently occurring features. This problem eludes to the credit assignment problem~\cite{ref:rumelhart:1986} common to these \ac{GD} variants.

      \item It is difficult to avoid getting trapped in local minima, especially for highly non-convex loss/cost functions used for \ac{ANN} training.~\cite{ref:dauphin:2014} argues that this difficulty arises not from local minima, but from sadle points. Sadle points are points at which one dimensions slopes upwards while another dimension slopes downwards.~\cite{ref:ruder:2016} mentions that these sadle points are usually surrounded by plateaus of the same error, leading to gradients that are close to zero in all dimensions.
\end{itemize}

@@

Research shows that \ac{SGD} has difficulty navigating ravines i.e. areas where the surface curves much more steeply in one dimension than in another~\cite{ref:sutton:1986}. These ravines are common around local minima. As such, \ac{SGD} is shown to oscillate across the slopes of the ravine while only making minor progress towards the local minima. An illustration of this is given in Figure~\ref{fig:heuristics:gd:sgd_with_and_without_momentum}.

\begin{figure}[htbp]
      \includegraphics[width=\textwidth]{images/sgd_with_and_without_momentum.png}
      \caption{An illustration of \ac{SGD} with and without momentum taken from~\cite{ref:du:2019}.}
      \label{fig:heuristics:gd:sgd_with_and_without_momentum}
\end{figure}

Momentum~\cite{ref:qian:1999} is a variant of the \acs{SGD} that helps accelerate \ac{SGD} in the relevant direction, dampening oscillations.~\cite{ref:ruder:2016} mentions that this is done by adding a fraction $\alpha$ of the weight update vector of the previous timestep to the current weight update vector. The accumulation of momentum is presented in Equation~\ref{eq:heuristics:gd:momentum_part_1} while the update step is then amended in Equation~\ref{eq:heuristics:gd:momentum_part_2} below.

@@

By redefining the \acs{SGD} update steps as shown above, the \acs{Momentum} heuristic allows for the increase of momentum for dimensions whose gradients point in the same direction, while simultaneously reducing momentum for dimensions whose gradients change direction, leading to faster convergence and less oscillation. The momentum term $\alpha$ is usually set to 0.9~\cite{ref:engelbrecht:2007}\cite{ref:ruder:2016}.

@@

\Acl{NAG} is a variant of the \acs{Momentum} heuristics developed by Ilya Sutskever~\cite{ref:sutskever:2013-2} and is inspired from Yurii Nesterov's~\cite{ref:nesterov:1983} work on optimising convex functions. \ac{NAG} provides an improvement to the momentum accumulation term by providing a look-ahead term that better refines the weight update step. In the \acs{NAG} heuristic, the gradient is not calculated w.r.t the current weights, but rather w.r.t the approximate future positions of the weights~\cite{ref:ruder:2016}. An illustration of this is taken from Geoffrey Hinton's lecture on mini-batch \ac{GD}~\cite{ref:hinton:2012} and is presented in Figure~\ref{fig:heuristics:gd:nag} below.

\begin{figure}[htbp]
      \includegraphics[width=\textwidth]{images/nag.png}
      \caption{An illustration of the weight update vector for \ac{NAG} taken from~\cite{ref:hinton:2012}.}
      \label{fig:heuristics:gd:nag}
\end{figure}

\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} explains the update step as is done by \ac{NAG} is follows. \ac{Momentum} first computes the current gradient. This is seen by the small blue vector presented in Figure~\ref{fig:heuristics:gd:nag}. A large step is taken in the direction of the updated accumulated gradient, presented by the big blue vector. \ac{NAG} then first takes a big step in the direction of the previous accumulated gradient presented by the brown vector. At this point the gradient is measured and \ac{NAG} then makes a correction (red vector), which results in the complete NAG update (green vector). By anticipating approximate future positions of the weights, the weight update step as defined by \ac{NAG} controls the optimisation process from from going too fast and results in increased responsiveness. The \ac{NAG} weight vector update steps are given in Equations~\ref{eq:heuristics:gd:nag_part_1} and~\ref{eq:heuristics:gd:nag_part_2} below.

@@

\Acl{Adagrad} is an adaptation of \ac{SGD} which implements a learning rate for every parameter in the weight vector and is developed by Duchi et al.~\cite{ref:duchi:2011}.~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that \Ac{Adagrad} adapts the learning rate to the weights, performing small updates (i.e. low learning rates) for weights associated with frequently occurring features and larger updates (i.e. high learning rates) for weights associated with infrequent features. For this reason \ac{Adagrad} is well suited for dealing with situations with very sparse training data.

In the \acs{GD} variants presented thus far, the weight updates have been applied to all weights at once using the same learning rate $\eta$ for all dimensions of the weight update vector. \Ac{Adagrad} uses a different learning rate for every weight $w_{i}$ at every timestep $t$. The weight update step as implemented by \ac{Adagrad} is presented in Equation~\ref{eq:heuristics:gd:adagrad_part_1} below.

@@

where $w_{i}$ is the $i$-th dimension of the weight vector, $G_{ii} \in \mathbb{R}^{d \times d}$ is a diagonal matrix where each diagonal element $i,i$ is the sum of the squared of the gradients w.r.t. $w_{i}$ up to timestep $t$ and $\epsilon$ is a smoothing term that avoids division by zero, usually set in the order of $1 \times 10^{-8}$~\cite{ref:ruder:2016}. Since $G_{t}$ contains the sum of the squares of the gradients with respect to all parameters $w$ along its diagonal, the weight update step can be vectorised and updated using the matrix-vector product between $G_{t}$ and $\nabla_{w_{i}}\mathcal{E}(w_{i})$. This is presented in Equation~\ref{eq:heuristics:gd:adagrad_part_2} below.

@@

In Section~\ref{sec:heuristics:adagrad} it was mentioned that one of the main problems of \ac{Adagrad} is that it has an aggressively monotonically decreasing learning rate because of the accumulation of the sum of squared gradients in its denominator. Zeiler~\cite{ref:zeiler:2012} presents an improvement on \ac{Adagrad} that eliminates this problem.~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that instead of accumulating all past squared gradients as in the case of \ac{Adagrad}, \ac{Adadelta} restricts the window of accumulation to a window with a fixed size $W$. However, storing $W$ previous squared gradients is very inefficient. Instead, the accumulation of the sum of gradients over $W$ steps is defined recursively as a decaying average of all past squared gradients. The moving average $E[g^{2}]_{t}$ at timestep $t$ then depends only a fraction $\alpha$ of the previous average and current gradient, similar to the update step for \ac{Momentum}. In order to simplify the notation of update steps, let $g = \nabla_{w_{i}}\mathcal{E}(w_{t})$. The moving average update step for $E[g^{2}]_{t}$ is then presented in Equation~\ref{eq:heuristics:gd:adadelta_part_1} below.

@@

Since the denominator is the \acs{RMS} error criterion of the gradient, the criteria short-hand notation is used as follows.

@@

\citeauthor{ref:ruder:2016}\cite{ref:ruder:2016} highlights that the original authors~\cite{ref:zeiler:2012} noted that the units of the update step as presented in Equation~\ref{eq:heuristics:gd:adadelta_part_2} do not match i.e. the update vector should have the same hypothetical units as the weight vector. To realize this, another exponentially decaying average is defined in terms of squared weight updates as is presented in Equation~\ref{eq:heuristics:gd:adadelta_part_6} below.
